Working with Daily Climate Model Output Data in R and the \pkg{futureheatwaves} Package

### 2017-05-24 paper_drafts/RJournalRevisions2/response_to_reviewers.txt

Added text: "While most CMIP5 climate models use the same calendar for all experiments, a few do not; a full table of the calendars used for each climate model and experiment, pulled from netCDF metadata, is available at https://www.earthsystemcog.org/projects/cog/faq_data."

In terms of the challenges presented by the large file sizes of climate
model output, we do recommend some pre-processing of data files to subset files down
to the geographic area and time range required for a study (e.g., North America for a national
study of the United States) which would help reduce data size prior to processing climate model
output through the described package. As the reviewer notes, we offer suggestions and guidance for
conducting this pre-processing in R, both through text in the manuscript. We offer additional
guidance through a vignette provided with the package described in the manuscript
(https://cran.r-project.org/web/packages/futureheatwaves/vignettes/starting_from_netcdf.html).
Even after pre-processing the files prior to using the `futureheatwaves` package, many research
applications will still involve working with very large files. For example, a national study of the
United States may still use files with daily temperatures for many decades at a large number of
grid points, and an analysis that uses multiple climate models, some with multiple ensemble
members, would still need to process a number of very large files. The `futureheatwaves` package
addresses these large file sizes by inputting and processing data from each ensemble member of each
climate model separately, avoiding the memory problems created by trying to load all data from all
ensemble members into memory at once and processing the data using `apply`, `map`, `plyr`, or
similar tools applied across all data at once.

We acknowledge, however, that while this approach is successful in terms of allowing the
identification and characterization of sustained extreme events in R, future software development
could continue refining and optimizing the handling of climate model data in R. We have added a
section to the end of the manuscript on future directions for climate model output data in R and
have included comments on how such future work could continue to tackle the challenges introduced
by the large file sizes of climate model output data:

Added text: "Another important path for development could be through approaches that allow
researchers to take advantage of the statistical tools offered by R while maintaining large climate
model output files in a netCDF format. For example, Goncalves and coauthors recently described a
"round table" approach of connecting as-needed data access from netCDF climate data files through
to functionality available in R and CDO through the intermediary of a MonetDB database system
(Goncalves et al., 2015). In other topical areas, R programmers are also improving the efficiency
of working with data in large netCDF files through approaches that avoid loading all data
in-memory. For example, the Bioconductor package `ncdfFlow` enables R users to conduct analyses of
hundreds of large flow cytometry output files through the creation and use of an `ncdfFlowSet`
class that stores the data in an HDF5 format rather than in-memory (Jiang et al., 2017;
Finak et al., 2014). Such an approach could be promising for future R software development for
working with large climate model files."

### 2017-05-24 paper_drafts/RJournalRevisions2/anderson-eason-barnes.Rmd
Continue work on revisions to R Journal manuscript

One important step in future development of R software to work with climate model output could be the development of R wrappers for some of the existing command line tools available through the Climate Data Operators (CDO) software \citep{schulzweida2006cdo}. Libraries already exist for Python and Ruby that allow the functionality of CDO tools to be used within these scripting languages (available from the Max Planck Institute at \url{https://code.zmaw.de/projects/cdo/wiki/Cdo%7Brbpy%7D}). While one R package (\CRANpkg{ncdf4.helpers}; \citet{ncdf.4helpers}) already provides R wrappers for a few CDO operators, such functionality could be extended through future R software to capture more of the full functionality of the CDO toolkit.

Another important path for development could be through approaches that allow researchers to take advantage of the statistical tools offered by R while maintaining large climate model output files in a netCDF format. For example, Goncalves and coauthors recently described a ``round table" approach of connecting as-needed data access from netCDF climate data files through to functionality available in R and CDO through the intermediary of a MonetDB database system \citep{goncalves2015round}. In other topical areas, R programmers are also improving the efficiency of working with data in large netCDF files through approaches that avoid loading all data in-memory. For example, the Bioconductor package \pkg{ncdfFlow} enables R users to conduct analyses of hundreds of large flow cytometry output files through the creation and use of an \code{ncdfFlowSet} class that stores the data in an HDF5 format rather than in-memory \citep{}. Such an approach could be promising for future R software development for working with large climate model files.

>>>>

One important step in future development of R software to work with climate model output could be the development of R wrappers for some of the existing command line tools available through the Climate Data Operators (CDO) software \citep{schulzweida2006cdo}. Libraries already exist for Python and Ruby that allow the functionality of CDO tools to be used within these scripting languages (available from the Max Planck Institute at \url{https://code.zmaw.de/projects/cdo/wiki/Cdo%7Brbpy%7D}). While one R package (\CRANpkg{ncdf4.helpers}; \citet{ncdf4helpers}) already provides R wrappers for a few CDO operators, such functionality could be extended through future R software to capture more of the full functionality of the CDO toolkit.

Another important path for development could be through approaches that allow researchers to take advantage of the statistical tools offered by R while maintaining large climate model output files in a netCDF format. For example, Goncalves and coauthors recently described a ``round table" approach of connecting as-needed data access from netCDF climate data files through to functionality available in R and CDO through the intermediary of a MonetDB database system \citep{goncalves2015round}. In other topical areas, R programmers are also improving the efficiency of working with data in large netCDF files through approaches that avoid loading all data in-memory. For example, the Bioconductor package \pkg{ncdfFlow} enables R users to conduct analyses of hundreds of large flow cytometry output files through the creation and use of an \code{ncdfFlowSet} class that stores the data in an HDF5 format rather than in-memory \citep{ncdfFlow, finak2014opencyto}. Such an approach could be promising for future R software development for working with large climate model files.

### 2017-05-24 paper_drafts/RJournalRevisions2/anderson-eason-barnes.tex
Continue work on revisions to R Journal manuscript

One important step in future development of R software to work with
climate model output could be the development of R wrappers for some of
the existing command line tools available through the Climate Data
Operators (CDO) software \citep{schulzweida2006cdo}. Libraries already
exist for Python and Ruby that allow the functionality of CDO tools to
be used within these scripting languages (available from the Max Planck
Institute at
\url{https://code.zmaw.de/projects/cdo/wiki/Cdo%7Brbpy%7D}). While one R
package (\CRANpkg{ncdf4.helpers}; \citet{ncdf.4helpers}) already
provides R wrappers for a few CDO operators, such functionality could be
extended through future R software to capture more of the full
functionality of the CDO toolkit.

Another important path for development could be through approaches that
allow researchers to take advantage of the statistical tools offered by
R while maintaining large climate model output files in a netCDF format.
For example, Goncalves and coauthors recently described a ``round table"
approach of connecting as-needed data access from netCDF climate data
files through to functionality available in R and CDO through the
intermediary of a MonetDB database system \citep{goncalves2015round}.

In other topical areas, R programmers are also improving the efficiency
of working with data in large netCDF files through approaches that work
with the data on-disc {[}?{]}. For example, the recent Bioconductor
package \pkg{openCyto} {[}?{]} allows scientists to work with the large
files generated by flow cytometry that characterize cell populations in
a sample. Similar approaches have been taken in metabolomics (e.g.,
{[}packages{]}) and other 'omics fields. The developers of future R
packages to work with climate model output may find it useful to borrow
ideas from these fields to create solutions with improved memory
management {[}?{]} for large climate output files.

>>>>

One important step in future development of R software to work with
climate model output could be the development of R wrappers for some of
the existing command line tools available through the Climate Data
Operators (CDO) software \citep{schulzweida2006cdo}. Libraries already
exist for Python and Ruby that allow the functionality of CDO tools to
be used within these scripting languages (available from the Max Planck
Institute at
\url{https://code.zmaw.de/projects/cdo/wiki/Cdo%7Brbpy%7D}). While one R
package (\CRANpkg{ncdf4.helpers}; \citet{ncdf4helpers}) already provides
R wrappers for a few CDO operators, such functionality could be extended
through future R software to capture more of the full functionality of
the CDO toolkit.

Another important path for development could be through approaches that
allow researchers to take advantage of the statistical tools offered by
R while maintaining large climate model output files in a netCDF format.
For example, Goncalves and coauthors recently described a ``round table"
approach of connecting as-needed data access from netCDF climate data
files through to functionality available in R and CDO through the
intermediary of a MonetDB database system \citep{goncalves2015round}. In
other topical areas, R programmers are also improving the efficiency of
working with data in large netCDF files through approaches that avoid
loading all data in-memory. For example, the Bioconductor package
\pkg{ncdfFlow} enables R users to conduct analyses of hundreds of large
flow cytometry output files through the creation and use of an
\code{ncdfFlowSet} class that stores the data in an HDF5 format rather
than in-memory \citep{ncdfFlow, finak2014opencyto}. Such an approach
could be promising for future R software development for working with
large climate model files.
